---
title: "Homework 2 - Into the Jungle"
author: "Summer 2016 - Balamuta"
date: "June 22, 2016"
layout: post
categories: hw
mathjax: true
tags: 
- debug
- recursion
- loops
- errors
- r
---

> "There are no routine statistical question; only questionable statistical routines." 
>
> --- **D.R. Cox** quoted in Chatfield, C. 1991. Avoiding statistical pitfalls. Statistical Science 6: 240-268.

# Directions

While working on homework, students are encouraged to work together in a group. Though, each student must turn in their own homework answers.  If students opt to work together, then it is important that they list the names of the students that they have collaborated with. Please note that sharing, copying or providing any part of a homework solution or code to another student is an infraction of the University’s rules on Academic Integrity. Rest assured, any violation of Academic Integrity will be punished **as severely as possible**.

* Your assignment must be submitted through the e-mail to `balamut2` [at] illinois [dot] edu. You are required to attach two (and only two) files to the *same* e-mail:
    * The RMarkdown (`.Rmd`) file must be saved as `hw2_yourNetID.Rmd`, e.g.  `hw2_balamut2.Rmd`
    * The generated output from the `.Rmd` file is then contained in `hw2_yourNetID.html`, e.g. `hw2_balamut2.Rmd`.
* Under this structure, the content found within the `.html` file will be considered a "report" which is the material that will determine the majority of your grade. Be sure to visibly include all `R` code and output that is relevant to answering the exercises. (You do not need to include irrelevant code you tried that resulted in error or did not answer the question correctly.)
* There are no limitations as to the number of submissions that can be made. However, only the last submission *before* the deadline will be graded by myself or the grader.

# Assignment Overview

**Deadline:** Sunday, June 26th by 11:59 PM CDT

## Objectives

1. Untangle and understand undocumented code
1. Document Code
1. Writing a Function
1. Logical Statements
1. Looping
1. Analzying the Big O of a function
1. Recursion
1. Memoziation Techniques

## Grading Criteria

We are going to be including a rubric that describes where each expectation lies. 

| Exercise | Exceptional | Exceeds Expectations |  Meets Expectations | Improvement Needed  | Unsatisfactory | Total Number of Points |
|----------|-------------|----------------------|---------------------|---------------------|----------------|------------------------|
|    0     |  $[0.9,1]$  | $[0.8,0.9)$          | $[0.7,0.8)$         |  $[0.6,0.7)$        | $[0,0.6)$      |           1            |
|    1     |  $[3.6,4]$  | $[3.2,3.6)$          | $[2.8,3.2)$         |  $[2.4,2.8)$        | $[0,2.4)$      |           4            |
|    2     |  $[4.5,5]$  | $[4,4.5)$            | $[3.5,4)$           |  $[3,3.5)$          | $[0,3)$        |           5            |
|    3     |  $[4.5,5]$  | $[4,4.5)$            | $[3.5,4)$           |  $[3,3.5)$          | $[0,3)$        |           5            |
|    4     |  $[4.5,5]$  | $[4,4.5)$            | $[3.5,4)$           |  $[3,3.5)$          | $[0,3)$        |           5            |
|    5     |  $[4.5,5]$  | $[4,4.5)$            | $[3.5,4)$           |  $[3,3.5)$          | $[0,3)$        |           5            |

# >>Assignment Rmd Document<<

[Feel free to use the assignment Rmd document instead of creating your own.](/assets/hw/hw2/hw2-assign.Rmd)

# Exercises

## (1 Point) Exercise 0 (Students To the GitHub Mobile!)

Keeping in line with our commitment to using versioning control, please:

1. Initialize a new repository on GitHub called `stat385hw2` 
2. Create a new RStudio project for Homework 2. 
3. **Please commit every exercise as you complete them.**
4. Provide a link to your GitHub repository to be awarded credit.

## (4 Points) Exercise 1 (It's a Mad, Mad, Mad, Mad Computational World)



The tenets behind this exercise are to prepare you to create different implementations and provide benchmarks for each implementation. 

Recall the different Sum of Squares, (*RSS*, *FSS*, *TSS*) formulations presented in [Lecture 5](http://stat385.thecoatlessprofessor.com/assets/lectures/lec5/lec5_functions_recursion_memoization_benchmarking.pdf#page=18). 

- Recreate the three sum of square functions given using loops **AND** recreate `tss_relationship` such that it uses these newly created functions.
   - Thus, you must "unvectorize" the computation. 
   - Call these functions: `compute_rss_loop`, `compute_fss_loop`, `compute_tss_loop`, and `tss_relationship_loop`.

- Check that the looping functions are equivalent to the vectorized components via `all.equal()`. Does the equality change between functions if the `tolerance` parameter in `all.equal` changes? (e.g. from `1e-3` to `0`) **Why?**
    - Use the [Setup Function Call](http://stat385.thecoatlessprofessor.com/assets/lectures/lec5/lec5_functions_recursion_memoization_benchmarking.pdf#page=22) to generate data.
    - Increase the number of observation in the script to `n = 1e4`
    - **Note:** You will have to use `check.attributes = F` to avoid `"names for current but not for target"`


- Perform the two different types of benchmarking operations using `rbenchmark` and `microbenchmark` as discussed in class. Comment as to whether there is a noticably difference in performance between the functions. Provide a recommendation as to the correct style to use in the future.
    - Make sure to **cache** the benchmark results in the `Rmd` document!


## (5 Points) Exercise 2 (Awake in a Nightmare)

The objective in this exercise is to understand unfamiliar code and improve it. While working on this problem, keep in mind the following quotation:

> "Always code as if the guy who ends up maintaining your code will be a violent psychopath who knows where you live. Code for readability."
> 
> --- John F. Woods ([Newsgroup comp.lang.c++](https://groups.google.com/forum/#!msg/comp.lang.c++/rYCO5yn4lXw/oITtSkZOtoUJ))


* Recently, a coworker left the company and your boss asked you to figure out the computations behind a graph the coworker had written. He wasn't entirely sure what was happening behind the scenes but the graph looked nice and the code could still be run. Your previous co-worker wrote the following code:

```{r, cache = T}
d = function(x){  length(x) }

s = function(x){ x*x }

u = function(x, ...){m = 0; k = 1; while(k <= d(x)){ m = m + x[k]; k = k + 1; }; m/d(x)}

w = function(b, v, q = T){n = d(b); o = 0; for(i in seq_len(n)){ o = o + s(b[i]-v) }; (1/(n-1*(q == T))*o)^(0.5)}

n = function(p, i, ...){ m = u(i, ...); r = w(i, m); (p-m)/r}

z = function(p, i, h = T, ...){
  g = seq(-3,3,length=1000);gg = dnorm(g)
  plot(g, gg, type="l", lwd=1, ...)
  k = n(p = p,i = i, ...)
  if(h) { tt = c(g[g<=k],k,-3); ll = c(gg[g<=k],0,0) }else{ tt = c(k,g[g>=k],3); ll = c(0,gg[g>=k],0) }
  polygon(tt, ll, col="red")
}
```

* The coworker left the following comment:

```{r coworker, cache = T, eval = F}
# Main function runs well
i = c(-1.21,0.28,1.08,-2.35,0.43,0.51,-0.57,-0.55,-0.56,-0.89)
z(1.57,i)
z(-2.33,i,h=F)
```

```{r, cache = T, echo = F}
<<coworker>>
```

1. First try to figure out what is happening 
    - What distribution does the graph look like?
    - What kind of calculation is occurring? 
    - Are there any redundancies in the code? (e.g. redefinition of common functions)
2. Document each part of the code using comments `# Function does...`
    - Provide clarity behind the parameters of the functions.
    - Emphasize what is being returned at each stage. 
    - You may need to reference help for function information via: `?`
3. Try to clean up the code. 
    - Use intuitive variable and function names
    - Correct the indentation and split statements up to multiple lines.
    - Implement vectorized or base function.
4. Make sure the output is the **same** between your version and the cleaned up version.
    - You may wish to individually test each function using `all.equal(bad_func(),your_func())`.
  
## (5 Points) Exercise 3 (Recursion is Recursion is Recursion is ...)

> "To iterate is human, to recurse divine."
>
> --- L. Peter Deutsch

Within this exercise, you will explore the concept of recursion and converting to and fro *tail* recursive statements. 

A common operation within Statistics is to obtain an **average** over data. The average in this case is given as: $${ {\bar x}_n} = \frac{1}{n}\sum\limits_{i = 1}^n { {x_i} }$$ where the subscript $n$ on ${ {\bar x} }$ represents the sample size.

- Implement two functions that obtains the average of a vector `x` where one function uses a loop and the other is vectorized.
    - The function must return `NaN` if the length of `x` is `0`. 
    - If the vector `x` contains `NA` values, then remove them using `na.omit()`.
    - **Note:** You are *not* permitted to call the base R implementation of `mean()`. However, you should check if the values are equal using `all.equal(x,y)`.

- We can also implement a recursive cummulative average if we modify the definition for $${ {\bar x}_n}$$ to depend on itself by ${ {\bar x}_{n-1} }$. 
    - Try to show that $${ {\bar x}_n} = \frac{1}{n}\left[ {\left( {n - 1} \right){ {\bar x}_{n - 1} } + {x_n} } \right]$$.

- Please implement the following recursive formula:
$${ {\bar x}_n} = \frac{1}{n}\left[ {\left( {n - 1} \right){ {\bar x}_{n - 1} } + {x_n} } \right]$$
    - Do not worry about returning `NaN` or detecting `NA` values. Focus only on writing the recursion case.

## (5 Points) Exercise 4 (Bugs, Bugs Everywhere!)

One of the many inveitiable side effects of programming is bugs. As Alexander so eloquently put it:

> "99 Little Bugs in the Code, take one down, patch it around, 117 little bugs in the code"
> 
> - Alexander via [Twitter](https://twitter.com/irqed/status/358212928404586498)

The next section contains many "errors" that appear in R code. These errors are actual cases that I have encountered while working with various students or on different projects. Your job is to solve them and explain *how* you fixed the problem. 

   1. Specify what the error is that you encounter when running the code.
   1. Briefly explain the implemented solution to fix the function.
   1. **One Caveat:** You are *not* allowed to modify the data that goes into the function. (Modifying the internals of functions is okay!)
   
- Why does my function not output anything!??!?
```{r infunc, eval = FALSE}
f = function(x){
  n = a+x
}

f(1)
```
- Huh, I could of sworn the function was working a minute ago. I made sure that the calculation were vectorized. Why doesn't it work!?
```{r, eval = FALSE}
p = function(x){
  log(p/(1-p))
}

vals = seq(0, 1, by = 0.1)

p(vals)
```
- The mean seems to be broken...
```{r, eval = FALSE}
d = c(1, 3, NA, 5, 6, 9)

mean(d) == 1/length(d)*sum(d)

```
- If this, then that right? 
```{r, eval = FALSE}
set.seed(145)

d = sample(1:10, 5)
b = sample(1:10, 5)

if(d > b){
  d+5
} 
else {
  b*2
}
```
- What is this blasphemy?! Why am I not getting any successful runs?? 
```{r, eval = FALSE}
set.seed(1234)

f = function() {
    if (runif(1) > .8) stop("oops")
    TRUE
}

g = function() {
  tryCatch(f(), 
           error = function(err) { warning(conditionMessage(err)) }
           )
}

a = replicate(10, g())
if(is.logical(a) && sum(a) > 0){
  cat("We had a few successes!\n")
}
```


## (5 Points) Exercise 5 (Memoize Me!)

The objective here is to construct different functions that are able to perform underneath memoization.

Recall from [Lecture 5](http://stat385.thecoatlessprofessor.com/assets/lectures/lec5/lec5_functions_recursion_memoization_benchmarking.pdf#page=37) the `factorial`.

- Create one *internal* and one *external* **memoized** factorial function.
  - Do **not** name it `factorial()`.
- Check for equality between the factorial functions using the base R implementation of `factorial()`.
- Perform a benchmark on all *four* functions. 
  - Compare the function against the base R and the two memoized versions.
- Use the function with the best *timing* to implement combinations: $$C\left( {n,k} \right) = \frac{ {n!} }{ {k!\left( {n - k} \right)!} }$$
  - Compare the combination function against `choose()` in base R. 
